---
title: "Getting started with bouquets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with bouquets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center"
)
library(bouquets)
set.seed(42)
```

## What is a bouquet plot?

A **bouquet plot** (also called an angular accumulation plot) converts the
directional dynamics of a time series into a 2-D turtle-graphics path:

- Each **increase** turns the heading left by a fixed angle $\theta$.
- Each **decrease** turns it right by $\theta$.
- No change keeps the heading straight.

All series share the same origin and angle, so paths that look alike come
from series with similar up/down patterns — not necessarily similar values.
The technique is structurally related to DNA walk visualisations (Gates 1986).

## Minimal example

```{r minimal}
set.seed(42)
n      <- 52L
weeks  <- seq(as.Date("2023-01-01"), by = "week", length.out = n)
season <- sin(seq(0, 2 * pi, length.out = n))

gw_long <- tibble::tibble(
  week    = rep(weeks, 3L),
  station = rep(c("Station A", "Station B", "Station C"), each = n),
  region  = rep(c("North", "North", "South"), each = n),
  level_m = c(
    8.5 + 0.8 * season + cumsum(rnorm(n,  0.00, 0.18)),
    7.2 + 0.5 * season + cumsum(rnorm(n,  0.02, 0.22)),
    9.1 + 1.1 * season + cumsum(rnorm(n, -0.01, 0.15))
  )
)

make_plot_bouquet(gw_long,
  time_col   = week,
  series_col = station,
  value_col  = level_m,
  verbose    = FALSE
)
```

Column order defaults make the call even shorter when your data is already in
the right column order: `make_plot_bouquet(gw_long, verbose = FALSE)`.

---

## Colour modes

Both `stem_colors` (the path lines) and `flower_colors` (the ✿ end markers)
accept four different types of input.

### 1 — Single hex colour

```{r single-hex}
make_plot_bouquet(gw_long,
  time_col      = week,
  series_col    = station,
  value_col     = level_m,
  stem_colors   = "#2d6a9f",
  flower_colors = "#e07b39",
  verbose       = FALSE
)
```

### 2 — Named keywords

`"greens"` and `"blossom"` select curated palettes:

```{r keywords}
make_plot_bouquet(gw_long,
  time_col      = week,
  series_col    = station,
  value_col     = level_m,
  stem_colors   = "greens",
  flower_colors = "blossom",
  verbose       = FALSE
)
```

### 3 — Vector of hex colours (one per series)

```{r vector}
make_plot_bouquet(gw_long,
  time_col      = week,
  series_col    = station,
  value_col     = level_m,
  stem_colors   = c("#c0392b", "#2980b9", "#27ae60"),
  verbose       = FALSE
)
```

### 4 — Column reference

Pass a bare column name. Each unique value in that column gets a distinct
perceptually uniform colour (via `hues::iwanthue()`):

```{r column-ref}
make_plot_bouquet(gw_long,
  time_col      = week,
  series_col    = station,
  value_col     = level_m,
  stem_colors   = region,
  flower_colors = region,
  verbose       = FALSE
)
```

---

## Faceting

`facet_by` accepts any column name and splits the plot via
`ggplot2::facet_wrap()`:

```{r facet}
make_plot_bouquet(gw_long,
  time_col      = week,
  series_col    = station,
  value_col     = level_m,
  stem_colors   = region,
  flower_colors = region,
  facet_by      = region,
  title         = "Groundwater dynamics by region",
  verbose       = FALSE
)
```

---

## Visual options

### Reference rings

`show_rings = TRUE` adds faint concentric circles at fixed radii to aid
distance judgements:

```{r rings}
make_plot_bouquet(gw_long,
  time_col   = week,
  series_col = station,
  value_col  = level_m,
  show_rings = TRUE,
  verbose    = FALSE
)
```

### Step markers

`marker_every` places a dot at every N-th step, useful for reading temporal
progress:

```{r markers}
make_plot_bouquet(gw_long,
  time_col     = week,
  series_col   = station,
  value_col    = level_m,
  marker_every = 13L,   # quarterly for weekly data
  verbose      = FALSE
)
```

### Dark mode

```{r dark}
make_plot_bouquet(gw_long,
  time_col      = week,
  series_col    = station,
  value_col     = level_m,
  stem_colors   = "greens",
  flower_colors = "blossom",
  dark_mode     = TRUE,
  title         = "Groundwater — dark mode",
  verbose       = FALSE
)
```

### Launch direction and ceiling percentage

`launch_deg` sets the initial heading (90° = upward, the default).
`ceiling_pct` controls how tight the turns are: values close to 1 make turns
pronounced; smaller values leave more headroom before a loop would occur.

```{r angle}
make_plot_bouquet(gw_long,
  time_col    = week,
  series_col  = station,
  value_col   = level_m,
  launch_deg  = 45,
  ceiling_pct = 0.60,
  verbose     = FALSE
)
```

---

## Clustering with `cluster_bouquet()`

`cluster_bouquet()` groups series by the similarity of their directional
sequences and appends a `cluster` factor column. The result plugs directly
into `make_plot_bouquet()`.

```{r cluster-basic}
set.seed(42)
n_big  <- 52L
weeks  <- seq(as.Date("2023-01-01"), by = "week", length.out = n_big)
season <- sin(seq(0, 2 * pi, length.out = n_big))

gw6 <- tibble::tibble(
  week    = rep(weeks, 6L),
  station = rep(paste0("S", 1:6), each = n_big),
  level_m = c(
    8.5 + 0.8 * season + cumsum(rnorm(n_big,  0.00, 0.2)),
    8.3 + 0.7 * season + cumsum(rnorm(n_big,  0.01, 0.2)),
    7.2 + 0.5 * season + cumsum(rnorm(n_big,  0.02, 0.2)),
    7.0 + 0.6 * season + cumsum(rnorm(n_big,  0.00, 0.2)),
    9.1 + 1.1 * season + cumsum(rnorm(n_big, -0.01, 0.2)),
    9.3 + 1.0 * season + cumsum(rnorm(n_big, -0.02, 0.2))
  )
)

clustered <- cluster_bouquet(gw6,
  time_col   = week,
  series_col = station,
  value_col  = level_m,
  verbose    = FALSE
)

dplyr::distinct(clustered, station, cluster)
```

Colour and facet the bouquet by cluster in one pipe:

```{r cluster-plot}
clustered |>
  make_plot_bouquet(
    time_col      = week,
    series_col    = station,
    value_col     = level_m,
    stem_colors   = cluster,
    flower_colors = cluster,
    facet_by      = cluster,
    title         = "Series grouped by directional dynamics",
    verbose       = FALSE
  )
```

### Clustering methods

Five algorithms are available via the `method` argument:

| Method | Description |
|---|---|
| `"pca_hclust"` | **(default)** PCA compression + Ward's D2 hclust. Best for long series. |
| `"pca_kmeans"` | PCA + k-means. Faster for many series. |
| `"hclust"` | Ward's D2 directly on direction sequences. Good for short series. |
| `"kmeans"` | k-means on raw sequences. |
| `"pam"` | Partitioning Around Medoids. Cluster centres are real observed series. |

The `distance` argument (`"euclidean"`, `"correlation"`, `"manhattan"`)
applies to `"hclust"` and `"pam"`. Correlation distance measures whether two
series **co-move** regardless of amplitude, which is often the most
meaningful choice for hydrological data:

```{r cluster-corr}
cluster_bouquet(gw6,
  time_col   = week,
  series_col = station,
  value_col  = level_m,
  method     = "hclust",
  distance   = "correlation",
  verbose    = FALSE
) |>
  dplyr::distinct(station, cluster)
```

### Automatic k selection

When `k = "auto"` (default), the function fits the model for
$k = 2, \ldots, k_\text{max}$ and selects the k that maximises:

$$\text{score}(k) = \bar{s}(k) \times \min_c \bar{s}_c(k) \times k^{\text{resolution}}$$

The worst-cluster term $\min_c \bar{s}_c(k)$ prevents choosing a k where
any one cluster is poorly defined. The `resolution` exponent rewards finer
groupings while they remain well-separated. Increase it to push toward more
clusters:

```{r cluster-resolution}
cluster_bouquet(gw6,
  time_col   = week,
  series_col = station,
  value_col  = level_m,
  resolution = 1.5,
  verbose    = FALSE
) |>
  dplyr::distinct(station, cluster)
```

Fix k directly when you already know how many groups you need:

```{r cluster-fixed-k}
cluster_bouquet(gw6,
  time_col   = week,
  series_col = station,
  value_col  = level_m,
  k          = 2L,
  verbose    = FALSE
) |>
  dplyr::distinct(station, cluster)
```

---

## Location map panel

When coordinate columns are supplied, `make_plot_bouquet()` attaches a
location map to the right of the bouquet. Points are coloured to match the
flower markers. The map background is drawn from the `maps` package (country
outlines) or `mapdata` (sub-national boundaries) if installed.

Coordinates in WGS84 decimal degrees work directly. For projected
coordinates supply the EPSG code via `coord_crs`:

```{r map, eval = requireNamespace("patchwork", quietly = TRUE) && requireNamespace("maps", quietly = TRUE)}
# Example with decimal-degree coordinates
gw_coords <- dplyr::mutate(gw_long,
  lon = c(rep(9.9, n), rep(13.4, n), rep(8.7, n)),
  lat = c(rep(51.5, n), rep(52.5, n), rep(50.1, n))
)

make_plot_bouquet(gw_coords,
  time_col   = week,
  series_col = station,
  value_col  = level_m,
  lon_col    = lon,
  lat_col    = lat,
  map_width  = 0.35,
  verbose    = FALSE
)
```

`map_width` controls the relative width of the map panel (default `0.35` =
35 % of the combined figure width). The map is automatically zoomed and
padded around the point locations, and works for any country or region.

---

## Combining everything

```{r full-pipeline, eval = FALSE}
  cluster_bouquet(
    time_col   = date,
    series_col = well_id,
    value_col  = gwl,
    method     = "pca_hclust",
    distance   = "euclidean",
    resolution = 0.8
  ) |>
  make_plot_bouquet(
    time_col      = date,
    series_col    = well_id,
    value_col     = gwl,
    stem_colors   = cluster,
    flower_colors = cluster,
    facet_by      = cluster,
    lon_col       = easting,
    lat_col       = northing,
    coord_crs     = 3035,
    show_rings    = TRUE,
    dark_mode     = TRUE,
    title         = "Groundwater dynamics by directional cluster",
    verbose       = FALSE
  )
```
