---
title: "Getting started with bouquets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with bouquets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(bouquets)
library(dplyr)
library(tibble)
```

<img src="../man/figures/logo.svg" align="right" height="139" alt="bouquets hex sticker" />

## What is a bouquet plot?

An angular accumulation plot encodes each time series as a turtle-graphics
path from a shared origin. At every time step:

- an **increase** turns the heading **left** by θ,
- a **decrease** turns it **right** by θ,
- **no change** goes straight ahead.

Because every series shares the same origin and angle, paths that look similar
belong to series with similar patterns of ups and downs — regardless of their
absolute values. The collection of paths radiates from the origin like stems in
a bouquet.

## Simulated data

Throughout this vignette we use a simulated set of eight groundwater monitoring
stations, each with a weekly time series containing a shared seasonal signal
plus independent noise.

```{r data}
set.seed(42)
n      <- 104L   # 2 years, weekly
weeks  <- seq(as.Date("2022-01-01"), by = "week", length.out = n)
season <- sin(seq(0, 4 * pi, length.out = n))

gw_long <- tibble::tibble(
  week    = rep(weeks, 8L),
  station = rep(paste0("S", 1:8), each = n),
  region  = rep(c("North", "North", "South", "South",
                  "East",  "East",  "West",  "West"), each = n),
  lon     = rep(c(7.2, 8.1, 13.4, 14.1, 10.5, 11.2, 6.9, 7.8), each = n),
  lat     = rep(c(53.6, 53.1, 52.5, 52.0, 50.8, 50.3, 48.1, 47.8), each = n),
  level_m = c(
    8.5 + 0.9 * season + cumsum(rnorm(n,  0.00, 0.18)),
    8.3 + 0.8 * season + cumsum(rnorm(n,  0.01, 0.20)),
    7.2 + 0.5 * season + cumsum(rnorm(n,  0.02, 0.22)),
    7.0 + 0.6 * season + cumsum(rnorm(n,  0.00, 0.19)),
    9.1 + 1.1 * season + cumsum(rnorm(n, -0.01, 0.15)),
    9.3 + 1.0 * season + cumsum(rnorm(n, -0.02, 0.16)),
    6.8 + 0.3 * season + cumsum(rnorm(n,  0.00, 0.28)),
    7.5 + 0.4 * season + cumsum(rnorm(n,  0.01, 0.25))
  )
)
```

## Minimal plot

The simplest call uses the first three columns as time, series, and value.

```{r minimal}
make_plot_bouquet(
  gw_long,
  time_col   = week,
  series_col = station,
  value_col  = level_m
)
```

## Colour and faceting

`stem_colors` and `flower_colors` each accept a single hex code, a vector of
hex codes, a keyword (`"greens"` / `"blossom"`), or a bare column name.

```{r colors-facets}
make_plot_bouquet(
  gw_long,
  time_col      = week,
  series_col    = station,
  value_col     = level_m,
  stem_colors   = region,
  flower_colors = region,
  facet_by      = region,
  title         = "Groundwater levels by region"
)
```

## Highlighting individual series

Use `highlight` to bring specific series to the foreground; all others are
dimmed to a pale grey.

```{r highlight}
make_plot_bouquet(
  gw_long,
  time_col   = week,
  series_col = station,
  value_col  = level_m,
  highlight  = c("S1", "S2"),
  title      = "Stations S1 and S2 highlighted"
)
```

## Time window filter

Use `from` and `to` to restrict the time range without pre-filtering.

```{r timewindow}
make_plot_bouquet(
  gw_long,
  time_col   = week,
  series_col = station,
  value_col  = level_m,
  from       = as.Date("2022-07-01"),
  to         = as.Date("2022-12-31"),
  title      = "Second half of 2022 only"
)
```

## Normalised mode

When `normalise = TRUE` each series receives its own per-series θ so every
path uses the full angular range independently of volatility. This aids shape
comparison across series with very different variances.

```{r normalise}
make_plot_bouquet(
  gw_long,
  time_col   = week,
  series_col = station,
  value_col  = level_m,
  normalise  = TRUE,
  title      = "Per-series normalised angles"
)
```

## Magnitude mode

`normalise = "magnitude"` replaces the binary ±1 step with the actual signed
delta value. The global calibration ensures the series with the widest value
range sweeps `ceiling_pct × 360°`, so large jumps bend the path more than
small ones. This encodes **both direction and amplitude** in the path shape,
making it possible to compare volatility across series: a volatile series
curves more than a flat one, and two series with similar variance but different
directional patterns produce visually distinct shapes.

```{r magnitude}
make_plot_bouquet(
  gw_long,
  time_col   = week,
  series_col = station,
  value_col  = level_m,
  normalise  = "magnitude",
  title      = "Magnitude mode -- turn angle proportional to delta"
)
```

## Clustering

`cluster_bouquet()` adds a `cluster` column to the data. It returns a
`cluster_bouquet` object: a normal tibble that also carries clustering
diagnostics. Call `summary()` to inspect them.

```{r cluster}
clustered <- cluster_bouquet(
  gw_long,
  time_col   = week,
  series_col = station,
  value_col  = level_m,
  seed       = 42L
)

summary(clustered)
```

### Visualise cluster quality

`plot_cluster_quality()` plots the composite silhouette score across all
candidate k values; the selected k is highlighted in pink.

```{r cluster-quality, fig.height=4}
plot_cluster_quality(clustered)
```

### Plot coloured and faceted by cluster

The `cluster` column works directly with `stem_colors`, `flower_colors`, and
`facet_by`.

```{r cluster-plot}
make_plot_bouquet(
  clustered,
  time_col      = week,
  series_col    = station,
  value_col     = level_m,
  stem_colors   = cluster,
  flower_colors = cluster,
  title         = "Stations coloured by directional cluster"
)
```

## Location map with cluster hulls

When `lon_col` and `lat_col` are supplied a location map is appended to the
right of the bouquet panel. Pass the cluster column to `cluster_hull` to draw
concave hulls around each cluster's points. Hull fill colour automatically
matches the cluster's `flower_colors`.

```{r map-hull, eval = requireNamespace("patchwork", quietly = TRUE) && requireNamespace("maps", quietly = TRUE)}
make_plot_bouquet(
  clustered,
  time_col      = week,
  series_col    = station,
  value_col     = level_m,
  stem_colors   = cluster,
  flower_colors = cluster,
  lon_col       = lon,
  lat_col       = lat,
  cluster_hull  = cluster,
  title         = "Colour-matched cluster hulls"
)
```

### Focusing on the cluster core with `hull_coverage`

When clusters are spatially dense or overlap, hulls drawn over all points can
obscure the map. `hull_coverage` (default `1`) restricts the hull to the
fraction of each cluster's points that lie closest to the cluster centroid.
Setting it to `0.75`, for example, excludes the outermost 25% of points before
computing the hull:

```{r map-hull-core, eval = requireNamespace("patchwork", quietly = TRUE) && requireNamespace("maps", quietly = TRUE)}
make_plot_bouquet(
  clustered,
  time_col      = week,
  series_col    = station,
  value_col     = level_m,
  stem_colors   = cluster,
  flower_colors = cluster,
  lon_col       = lon,
  lat_col       = lat,
  cluster_hull  = cluster,
  hull_coverage = 0.75,   # hull covers the innermost 75% of each cluster
  title         = "Cluster hulls: core 75%"
)
```

## Saving plots

`save_bouquet()` wraps `ggplot2::ggsave()` with sensible defaults for the
bouquet aspect ratio.

```{r save, eval=FALSE}
p <- make_plot_bouquet(gw_long, week, station, level_m)
save_bouquet(p, "my_bouquet.png", width = 10, height = 8)
```

## Interactive version

`make_plot_bouquet_interactive()` produces a plotly figure with hover tooltips.
Requires the `plotly` package.

```{r interactive, eval=FALSE}
make_plot_bouquet_interactive(gw_long, week, station, level_m)
```

## Full pipeline

A compact end-to-end workflow:

```{r pipeline, eval=FALSE}
gw_long |>
  cluster_bouquet(
    time_col   = week,
    series_col = station,
    value_col  = level_m,
    seed       = 42L
  ) |>
  make_plot_bouquet(
    time_col      = week,
    series_col    = station,
    value_col     = level_m,
    stem_colors   = cluster,
    flower_colors = cluster,
    lon_col       = lon,
    lat_col       = lat,
    cluster_hull  = cluster,
    hull_coverage = 0.75,
    title         = "Groundwater directional dynamics"
  )
```
